---
title: "Alzheimer's Simulation Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a data simulation project based of a study conducted by Friedland et al that was published in 2001 in PNAS called: 

Patients with Alzheimer's disease have reduced activities in midlife compared with healthy control-group members
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC30672/

It was a study that was focused on the correlations between activity levels and Alzheimer's disease (AD). Previous studies have shown certain activities to be correlated with a higher risk for AD. In this study, the focus is instead on identifying correlations between activity levels and AD, which showed that reduced activity levels correlated with AD. 

The simulated data is based on a survey of different activities that fall under three types:

a) passive activities (ex. watching TV)

b) intellectual activities (ex. reading a book)

c) physical activities (ex. jogging)


The primary predictor variables are:

1. Diversity - How many different activities within each type (passive, intellectual, physical)?

2. Intensity - How many hours per month invested per activity?

3. Percent Intensity - Fraction of hours invested in each type of activity

4. Age - Assessed activities during early adulthood (20-30s) and middle adulthood (40-50s)


STEP 1: Loading required packages and setting seed
```{r}
library(dplyr)
library(corrplot)
library(ggplot2)
set.seed(100)
```

STEP 2: Generating participant data
```{r}
#generate age data for AD cases
age.m <- data.frame(rnorm(250, 72.2, 7.3))
age.f <- data.frame(rnorm(250, 72.7, 8.5))
names(age.m) <- "age"
names(age.f) <- "age"
summary(age.m) #summary of age for male AD cases
summary(age.f) #summary of age for female AD cases
age.m$gender <- "Male"
age.f$gender <- "Female"
dat1 <- rbind(age.m, age.f)
#sample age data
dat1 <- sample_n(dat1, 193)
dat1$category <- "AD"

##generate age data for AD controls
age.mc <- data.frame(rnorm(500, 71.7, 5.2))
age.fc <- data.frame(rnorm(500, 71, 6.4))
names(age.mc) <- "age"
names(age.fc) <- "age"
summary(age.mc) #summary of age for male control cases
summary(age.fc) #summary of age for female control cases
age.mc$gender <- "Male"
age.fc$gender <- "Female"
dat2 <- rbind(age.mc, age.fc)
#sample age data
dat2 <- sample_n(dat2, 358)
dat2$category <- "Control"
```

STEP 3: Simulate data for AD cases based on reported mean and SD
```{r}
n <- 193
#diversity
pass.div <- rnorm(n, 0.84, 0.18)
int.div <- rnorm(n, 0.43, 0.19)
phys.div <- rnorm(n, 0.39, 0.22)
#replace negative values with zero
pass.div <- ifelse(pass.div<0, 0, pass.div)
int.div <- ifelse(int.div<0, 0, int.div)
phys.div <- ifelse(phys.div<0, 0, phys.div)

#early adulthood intensity
pass.itsE <- rnorm(n, 59.28, 50.27)
int.itsE <- rnorm(n, 55.63, 40.36)
phys.itsE <- rnorm(n, 32.06, 33.66)
#replace negative values with zero
pass.itsE <- ifelse(pass.itsE<0, 0, pass.itsE)
int.itsE <- ifelse(int.itsE<0, 0, int.itsE)
phys.itsE <- ifelse(phys.itsE<0, 0, phys.itsE)
#replace high numbers with mean + 1 SD
pass.itsE <- ifelse(pass.itsE>59.28+50.27, 59.28, pass.itsE)
int.itsE <- ifelse(int.itsE>55.63+40.36, 55.63, int.itsE)
phys.itsE <- ifelse(phys.itsE>32.06+33.66, 32.06, phys.itsE)
```

STEP 4: Correlate middle adulthood intensity to early adulthood intensity for AD cases
```{r}
#middle adulthood intensity - adjusted for Alzheimer's category using early adulthood intensity
pass.itsM <- pass.itsE*rnorm(n, (98.9/59.28), ((68.82/98.9)+(50.27/59.28))/4)
int.itsM <- int.itsE*rnorm(n, (68.06/55.63), ((59.8/68.06)+(40.36/55.63))/4)
phys.itsM <- phys.itsE*rnorm(n, (38.12/32.06), ((38.13/38.12)+(33.66/32.06))/4)
#replace negative values with zero
pass.itsM <- ifelse(pass.itsM<0, 0, pass.itsM)
int.itsM <- ifelse(int.itsM<0, 0, int.itsM)
phys.itsM <- ifelse(phys.itsM<0, 0, phys.itsM)
#replace high numbers with mean + 1 SD
pass.itsM <- ifelse(pass.itsM>98.9+68.82, 98.9, pass.itsM)
int.itsM <- ifelse(int.itsM>68.06+59.8, 68.06, int.itsM)
phys.itsM <- ifelse(phys.itsM>38.12+38.13, 38.12, phys.itsM)

#cbind all AD features
dat1 <- cbind(pass.div, int.div, phys.div, pass.itsE, int.itsE, phys.itsE,
              pass.itsM, int.itsM, phys.itsM, dat1)
```

STEP 5: Simulate data for control cases based on reported mean and SD
```{r}
n <- 358
#diversity
pass.div <- rnorm(n, 0.91, 0.14)
int.div <- rnorm(n, 0.54, 0.16)
phys.div <- rnorm(n, 0.55, 0.22)
#replace negative values with zero
pass.div <- ifelse(pass.div<0, 0, pass.div)
int.div <- ifelse(int.div<0, 0, int.div)
phys.div <- ifelse(phys.div<0, 0, phys.div)

#early adulthood intensity
pass.itsE <- rnorm(n, 69.88, 55.71)
int.itsE <- rnorm(n, 68.79, 52.23)
phys.itsE <- rnorm(n, 33.97, 34.58)
#replace negative values with zero
pass.itsE <- ifelse(pass.itsE<0, 0, pass.itsE)
int.itsE <- ifelse(int.itsE<0, 0, int.itsE)
phys.itsE <- ifelse(phys.itsE<0, 0, phys.itsE)
#replace high numbers with mean + 1 SD
pass.itsE <- ifelse(pass.itsE>69.88+55.71, 69.88, pass.itsE)
int.itsE <- ifelse(int.itsE>68.79+52.23, 68.79, int.itsE)
phys.itsE <- ifelse(phys.itsE>34.58+33.97, 33.97, phys.itsE)
```

STEP 6: Correlate middle adulthood intensity to early adulthood intensity for control cases
```{r}
#middle adulthood intensity
pass.itsM <- pass.itsE*rnorm(n, (101.89/69.88), ((61.31/101.89)+(55.71/69.88))/4)
int.itsM <- int.itsE*rnorm(n, (79.14/68.79), ((55.66/79.14)+(52.23/68.79))/4)
phys.itsM <- phys.itsE*rnorm(n, (40.8/33.97), ((33.43/40.8)+(34.58/33.97))/4)
#replace negative values with zero
pass.itsM <- ifelse(pass.itsM<0, 0, pass.itsM)
int.itsM <- ifelse(int.itsM<0, 0, int.itsM)
phys.itsM <- ifelse(phys.itsM<0, 0, phys.itsM)
#replace high numbers with mean + 1 SD
pass.itsM <- ifelse(pass.itsM>101.89+61.31, 101.89, pass.itsM)
int.itsM <- ifelse(int.itsM>79.14+55.66, 79.14, int.itsM)
phys.itsM <- ifelse(phys.itsM>40.8+33.43, 40.8, phys.itsM)

#cbind all control features
dat2 <- cbind(pass.div, int.div, phys.div, pass.itsE, int.itsE, phys.itsE,
              pass.itsM, int.itsM, phys.itsM, dat2)

#rbind all features
dats <- rbind(dat1, dat2)
```

STEP 7: Clean data based on inconsistencies (ex. non-zero intensity for zero diversity - 20hrs for 0 passive activities) 
```{r}
#replace all intensity/percent intensity with zeroes for zero diversity
dats$pass.itsE <- ifelse(dats$pass.div==0, 0, dats$pass.itsE)
dats$int.itsE <- ifelse(dats$int.div==0, 0, dats$int.itsE)
dats$phys.itsE <- ifelse(dats$phys.div==0, 0, dats$phys.itsE)
dats$pass.itsM <- ifelse(dats$pass.div==0, 0, dats$pass.itsM)
dats$int.itsM <- ifelse(dats$int.div==0, 0, dats$int.itsM)
dats$phys.itsM <- ifelse(dats$phys.div==0, 0, dats$phys.itsM)

#replace all zero intensity for non-zero diversity (0 hrs for 2 passive activities)
#pass intensity early
for (k in seq(nrow(dats))){
  if (dats$pass.div[k]!=0 && dats$pass.itsE[k]==0){
    while (dats$pass.itsE[k]==0){
      if (dats$category[k]=="AD"){
        dats$pass.itsE[k] <- rnorm(1, 59.28, 50.27)
        dats$pass.itsE[k] <- ifelse(dats$pass.itsE[k]<0, 0, dats$pass.itsE[k])
      } else {
        dats$pass.itsE[k] <- rnorm(1, 69.88, 55.71)
        dats$pass.itsE[k] <- ifelse(dats$pass.itsE[k]<0, 0, dats$pass.itsE[k])
      }
    }
  }
}

#int intensity early
for (k in seq(nrow(dats))){
  if (dats$int.div[k]!=0 && dats$int.itsE[k]==0){
    while (dats$int.itsE[k]==0){
      if (dats$category[k]=="AD"){
        dats$int.itsE[k] <- rnorm(1, 55.63, 40.36)
        dats$int.itsE[k] <- ifelse(dats$int.itsE[k]<0, 0, dats$int.itsE[k])
      } else {
        dats$int.itsE[k] <- rnorm(1, 68.79, 52.23)
        dats$int.itsE[k] <- ifelse(dats$int.itsE[k]<0, 0, dats$int.itsE[k])
      }
    }
  }
}

#phys intensity early
for (k in seq(nrow(dats))){
  if (dats$phys.div[k]!=0 && dats$phys.itsE[k]==0){
    while (dats$phys.itsE[k]==0){
      if (dats$category[k]=="AD"){
        dats$phys.itsE[k] <- rnorm(1, 32.06, 33.66)
        dats$phys.itsE[k] <- ifelse(dats$phys.itsE[k]<0, 0, dats$phys.itsE[k])
      } else {
        dats$phys.itsE[k] <- rnorm(1, 33.97, 34.58)
        dats$phys.itsE[k] <- ifelse(dats$phys.itsE[k]<0, 0, dats$phys.itsE[k])
      }
    }
  }
}

#pass intensity middle
for (k in seq(nrow(dats))){
  if (dats$pass.div[k]!=0 && dats$pass.itsM[k]==0){
    while (dats$pass.itsM[k]==0){
      if (dats$category[k]=="AD"){
        dats$pass.itsM[k] <- rnorm(1, 98.9, 68.82)
        dats$pass.itsM[k] <- ifelse(dats$pass.itsM[k]<0, 0, dats$pass.itsM[k])
      } else {
        dats$pass.itsM[k] <- rnorm(1, 101.89, 61.31)
        dats$pass.itsM[k] <- ifelse(dats$pass.itsM[k]<0, 0, dats$pass.itsM[k])
      }
    }
  }
}

#int intensity middle
for (k in seq(nrow(dats))){
  if (dats$int.div[k]!=0 && dats$int.itsM[k]==0){
    while (dats$int.itsM[k]==0){
      if (dats$category[k]=="AD"){
        dats$int.itsM[k] <- rnorm(1, 68.06, 59.8)
        dats$int.itsM[k] <- ifelse(dats$int.itsM[k]<0, 0, dats$int.itsM[k])
      } else {
        dats$int.itsM[k] <- rnorm(1, 79.14, 55.66)
        dats$int.itsM[k] <- ifelse(dats$int.itsM[k]<0, 0, dats$int.itsM[k])
      }
    }
  }
}

#phys intensity middle
for (k in seq(nrow(dats))){
  if (dats$phys.div[k]!=0 && dats$phys.itsM[k]==0){
    while (dats$phys.itsM[k]==0){
      if (dats$category[k]=="AD"){
        dats$phys.itsM[k] <- rnorm(1, 38.12, 38.13)
        dats$phys.itsM[k] <- ifelse(dats$phys.itsM[k]<0, 0, dats$phys.itsM[k])
      } else {
        dats$phys.itsM[k] <- rnorm(1, 40.8, 33.43)
        dats$phys.itsM[k] <- ifelse(dats$phys.itsM[k]<0, 0, dats$phys.itsM[k])
      }
    }
  }
}

#sum of all intensities to check for unrealistic values
dats$pass.itstotE <- dats$pass.itsE + dats$int.itsE + dats$phys.itsE
dats$pass.itstotM <- dats$pass.itsM + dats$int.itsM + dats$phys.itsM
```

STEP 8: Correlate percent intensity to simulated intensity data
```{r}
#early adult percent intensity
dats$pass.perE <- 100*dats$pass.itsE / (dats$pass.itsE+dats$int.itsE+dats$phys.itsE)
dats$int.perE <- 100*dats$int.itsE / (dats$pass.itsE+dats$int.itsE+dats$phys.itsE)
dats$phys.perE <- 100*dats$phys.itsE / (dats$pass.itsE+dats$int.itsE+dats$phys.itsE)

#middle adult percent intensity
dats$pass.perM <- 100*dats$pass.itsM / (dats$pass.itsM+dats$int.itsM+dats$phys.itsM)
dats$int.perM <- 100*dats$int.itsM / (dats$pass.itsM+dats$int.itsM+dats$phys.itsM)
dats$phys.perM <- 100*dats$phys.itsM / (dats$pass.itsM+dats$int.itsM+dats$phys.itsM)

#pass percent difference
dats$passChg <- dats$pass.perM - dats$pass.perE
dats$intChg <- dats$int.perM - dats$int.perE
dats$physChg <- dats$phys.perM - dats$phys.perE

#reorganize columns
dats <- dats[, c(1:9, 13:23, 10:12)]
dats$gender <- factor(dats$gender)
dats$category <- factor(dats$category)

#summary
summary(dats)
```

STEP 9: Visualize simulated data
```{r}
#correlation plots
c <- cor(dats[,c(1:18)], use= "pairwise.complete.obs")
corrplot(c)

#scatter plots of correlated variables
g <- ggplot(dats)
#correlated data between intensity and percent intensity
g+geom_point(aes(x=int.itsE, y=int.perE, color=category))
#correlated data between early adulthood intensity and middle adulthood intensity
g+geom_point(aes(x=int.itsE, y=int.itsM, color=category))
#comparison of diversity between AD and control cases
g+geom_boxplot(aes(x=category, y=int.div, fill=category))
#comparison of diversity by density plot
g+geom_density(aes(x=int.div, fill=category, alpha=0.3))
#comparison of percent intensity changes from early to middle adulthood
g+geom_point(aes(x=int.itsE, y=intChg, color=category))
#comparison of percent intensity between AD and control cases
g+geom_boxplot(aes(x=category, y=intChg, fill=category))
```

STEP 10: Assess mean differences in simulated intensity and percent intensity data
```{r}
#separate by category
subAlz <- filter(dats, category=="AD")
subCtr <- subset(dats, category=="Control")

#comparison of early adulthood intensity
t.test(subAlz$int.itsE, subCtr$int.itsE)
#comparison of early adulthood percent intensity
t.test(subAlz$int.perE, subCtr$int.perE)
#comparison of middle adulthood intensity
t.test(subAlz$int.itsM, subCtr$int.itsM)
#comparison of middle adulthood percent intensity
t.test(subAlz$int.perM, subCtr$int.perM)
#comparison of changes in percent intensity between AD and control cases
t.test(subAlz$intChg, subCtr$intChg)
```

CONCLUSIONS:
The primary conclusions of the study was that reduced activity levels correlated with higher risk for AD. Based on the simulated data, diversity measurements were reduced for AD cases across all three types of activities (passive, intellectual, and physical). Additionally, the simulated data indicates a significant difference in the intensity and percent intensity levels between AD and control cases. However, the changes in the percent intensities between AD and control cases was insignificant. This may be due to the absurdly high report values for the standard deviations. 